---
title: "Cyclistic Data Analysis"
author: "Patricia Ellis"
date: "08/29/2021"
output:
  html_document:
    df_print: paged
---

 

### Deliverables:
Produce a report with the following deliverables:   
1. A clear statement of the business task   
2. A description of all data sources used   
3. Documentation of any cleaning or manipulation of data   
4. A summary of your analysis   
5. Supporting visualizations and key findings   
6. Your top three recommendations based on your analysis  


### Business Task:
How do annual members and casual riders use Cyclistic bikes differently?

### Data Source:
Motivate International Inc. Public use
12 months of divvy trip data from 8/2020 to 7/2021
Saved in local folder, uploaded to Google Cloud bucket and uploaded to Big Query

I will be working in R studio due to the large amount of data.

Load packages
```{r include=FALSE}
install.packages("tidyverse")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("lubridate")
install.packages("stringr")
install.packages("purrr")
install.packages("forcats")
install.packages("geosphere")
install.packages("gridExtra")
install.packages("data.table")
install.packages("ggmap")

```


Load the libraries
```{r include=FALSE}
library(tidyverse)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(stringr)
library(forcats)
library(purrr)
library(geosphere)
library(gridExtra)
library(data.table)
library(ggmap)

```

Install and read the data sets.

```{r}
Aug_20 <- read_csv("202008-divvy-tripdata.csv")
Sep_20 <- read_csv("202009-divvy-tripdata.csv")
Oct_20 <- read_csv("202010-divvy-tripdata.csv")
Nov_20 <- read_csv("202011-divvy-tripdata.csv")
Dec_20 <- read_csv("202012-divvy-tripdata.csv")
Jan_21 <- read_csv("202101-divvy-tripdata.csv")
Feb_21 <- read_csv("202102-divvy-tripdata.csv")
Mar_21 <- read_csv("202103-divvy-tripdata.csv")
Apr_21 <- read_csv("202104-divvy-tripdata.csv")
May_21 <- read_csv("202105-divvy-tripdata.csv")
Jun_21 <- read_csv("202106-divvy-tripdata.csv")
Jul_21 <- read_csv("202107-divvy-tripdata.csv")
```

Some data types for start and end_station need to be changed.

```{r}
Nov_20 <- mutate(Nov_20, start_station_id = as.character(start_station_id), 
                   end_station_id = as.character(end_station_id))
Oct_20 <- mutate(Oct_20, start_station_id = as.character(start_station_id), 
                   end_station_id = as.character(end_station_id))
Sep_20 <- mutate(Sep_20, start_station_id = as.character(start_station_id), 
                   end_station_id = as.character(end_station_id))
Aug_20 <- mutate(Aug_20, start_station_id = as.character(start_station_id), 
                   end_station_id = as.character(end_station_id))

```

Lets check that it worked

```{r}
str(Nov_20)
```
Now we want to combine our data into a single table

```{r}
total_year <- bind_rows(Aug_20, Sep_20, Oct_20, Nov_20, Dec_20, Jan_21, Feb_21, Mar_21, Apr_21, May_21, Jun_21, Jul_21)

```

### Cleaning:
Check for duplicates in ride_id

```{r}
total_year <- total_year[!duplicated(total_year$ride_id), ]
print(paste("Removed", nrow(total_year) - nrow(total_year), "duplicated rows"))
```



```{r}
is.na_remove <- total_year$ride_id[!is.na(total_year$ride_id)]
```

```{r}
is.na_remove <- total_year$rideable_type[!is.na(total_year$rideable_type)]
```

```{r}
is.na_remove <- total_year$started_at[!is.na(total_year$started_at)]
```

```{r}
is.na_remove <- total_year$ended_at[!is.na(total_year$ended_at)]
```

```{r}
is.na_remove <- total_year$start_station_name[!is.na(total_year$start_station_name)]
```

```{r}
is.na_remove <- total_year$start_station_id[!is.na(total_year$start_station_id)]
```

```{r}
is.na_remove <- total_year$end_station_name[!is.na(total_year$end_station_name)]
```

```{r}
is.na_remove <- total_year$end_station_id[!is.na(total_year$end_station_id)]
```

```{r}
is.na_remove <- total_year$start_lat[!is.na(total_year$start_lat)]
```

```{r}
is.na_remove <- total_year$start_lng[!is.na(total_year$start_lng)]
```

```{r}
is.na_remove <- total_year$end_lat[!is.na(total_year$end_lat)]
```

```{r}
is.na_remove <- total_year$end_lng[!is.na(total_year$end_lng)]
```

```{r}
is.na_remove <- total_year$member_casual[!is.na(total_year$member_casual)]
```

### Analysis:

Let's calculate ride duration and create a new column.

```{r}
total_year$ride_duration <-
  difftime(total_year$ended_at, total_year$started_at, units = "mins")
```

```{r}
is.na_remove <- total_year$ride_duration[!is.na(total_year$ride_duration)]
```






Let's determine the day of the week for each ride.

```{r}
total_year$day_of_week <-
  as.integer(wday(as.Date(total_year$started_at), week_start = 1))
```


Calculate the distance:

```{r}
total_year$ride_distance <- distGeo(matrix(c(total_year$start_lng, total_year$start_lat), ncol = 2), matrix(c(total_year$end_lng, total_year$end_lat), ncol = 2))
total_year$ride_distance <- total_year$ride_distance/1000

```

Checking for trip duration less than 0

```{r}
nrow(subset(total_year,ride_duration < 0))
```
Remove the negative trip durations   
```{r}
total_year <- total_year[!(total_year$ride_duration < 0),]
```

Checking for test rides that were made by company for quality checks  
```{r}
nrow(subset(total_year, start_station_name %like% "TEST"))
nrow(subset(total_year, start_station_name %like% "test"))
nrow(subset(total_year, start_station_name %like% "Test"))
```
Remove test rides  
```{r}
total_year<- total_year[!((total_year$start_station_name %like% "TEST" | total_year$start_station_name %like% "test")),]
```

Check the dataframe  
```{r}
glimpse(total_year)
```
### Analysis  
Aggregating total ride duration by user type
```{r}
setNames(aggregate(ride_duration ~ member_casual, total_year, sum), c("userType", "total_ride_duration(mins)"))
```




Let's look at the number of rides per rider type.

```{r}
total_year %>% count(member_casual)
```
Who takes more rides, members or casual users?  
Members have taken 526,923 more rides than casual users.

What are the average duration and distances for user type?

Calculate the averages:

```{r}
is.na_remove <- total_year$ride_distance[!is.na(total_year$ride_distance)]
```

```{r echo=TRUE}
userType_mean <- total_year %>% group_by(member_casual) %>% summarise(mean_time = mean(ride_duration), mean_distance = mean(ride_distance, na.rm = TRUE))

membervstime <- ggplot(userType_mean) + 
                geom_col(mapping=aes(x=member_casual,y=mean_time,fill=member_casual), show.legend = FALSE) +
                labs(title = "Mean Travel Time by User Type",x="User Type",y="Mean Time in Sec")

membervsdistance <- ggplot(userType_mean) + 
                    geom_col(mapping=aes(x=member_casual,y=mean_distance,fill=member_casual), show.legend = FALSE)+
                    labs(title = "Mean Travel Distance by User Type",x="User Type",y="Mean distance In km",caption = "Data by Motivate International Inc")

grid.arrange(membervstime, membervsdistance, ncol = 2, nrow = 1)
```

Let's Look at the days of the week.


```{r}
total_year %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_duration),.groups = 'drop') %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of Rides by User Type by Week Day",x="Days of the week",y="Number of rides",caption = "Data by Motivate International Inc", fill="User type") +
  theme(legend.position="top")
```

Who takes longer duration rides, members or casual users?  
Casual users spend more time on the bikes, it's a good assumption that members are using the bikes to commute to work and don't waste time in getting there.

What day(s) of the week have the most riders by user type?  
Casual users are riding on the weekends more than week days suggesting that they may be tourists visiting on the weekend.  
Members are more prevalent on the week days further suggesting the idea that they are commuting to work.

Average ride time by each day for members vs casual users  
```{r include=FALSE}
aggregate(total_year$ride_duration ~ total_year$member_casual + total_year$day_of_week, FUN = mean)
```

```{r}
total_year %>% 
  mutate(weekday = day_of_week) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_duration)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```
Which bike type is the most popular?

```{r}
total_year %>%
  group_by(member_casual,rideable_type) %>%
  summarise(number_of_rides = n(),rideable_type) %>%
  arrange(member_casual, rideable_type) %>%
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

Does station location have a correlation to user type?  
#Create a table only for the most popular routes (>250 times)
```{r}
coordinates_table <- total_year %>% 
filter(start_lng != end_lng & start_lat != end_lat) %>%
group_by(start_lng, start_lat, end_lng, end_lat, member_casual, rideable_type) %>%
summarise(total = n(),.groups="drop") %>%
filter(total > 250)
```

#Create two sub tables for each user type  
```{r}
casual <- coordinates_table %>% filter(member_casual == "casual")
member <- coordinates_table %>% filter(member_casual == "member")
```

#Store the bounding box coordinates for ggmap:

```{r}
chi_bb <- c(
  left = -87.700424,
  bottom = 41.790769,
  right = -87.554855,
  top = 41.990119
)

#Store the stamen map of Chicago  

chicago_stamen <- get_stamenmap(
  bbox = chi_bb,
  zoom = 14,
  maptype = "toner"
)

```

```{r}
ggmap(chicago_stamen,darken = c(0.8, "white")) +
   geom_curve(casual, mapping = aes(x = start_lng, y = start_lat, xend = end_lng, yend = end_lat, alpha= total, color= rideable_type), size = 0.5, curvature = .2,arrow = arrow(length=unit(0.2,"cm"), ends="first", type = "closed")) +
    coord_cartesian() +
    labs(title = "Most popular routes by casual users",x=NULL,y=NULL, color="User type", caption = "Thanks to Julen Aranguren for the notebook on Kaggle, Map tiles by Stamen Design, under CC BY 3.0. Data by OpenStreetMap, under CC BY SA and ggmap") +
    theme(legend.position="bottom")

ggmap(chicago_stamen,darken = c(0.8, "white")) +
    geom_curve(member, mapping = aes(x = start_lng, y = start_lat, xend = end_lng, yend = end_lat, alpha= total, color=rideable_type), size = 0.5, curvature = .2,arrow = arrow(length=unit(0.2,"cm"), ends="first", type = "closed")) +  
    coord_cartesian() +
    labs(title = "Most popular routes by annual members",x=NULL,y=NULL, caption = "Data by Motivate International Inc Thanks to Julen Aranguren for the maps on Kaggle
Map tiles by Stamen Design, under CC BY 3.0. Data by OpenStreetMap, under CC BY SA. ggmap") +
    theme(legend.position="bottom")

```

### Recommendations


